\section{Algorithms}
\label{app:algorithms}

One of the most popular algorithms to generate realizations of the
CTMC in the well-stirred case is Gillespie's direct method (DM)
\cite{SSA}. Several algorithmic improvements of this method exist,
one of them being the next reaction method (NRM) due to Gibson and
Bruck \cite{NRM}.

The underlying algorithm in URDME is the next subvolume method (NSM)
\cite{BISTAB}. The NSM can be understood as a combination of NRM and
DM in order to tailor the algorithm to reaction-diffusion processes.

For reference, we first state below both DM and NRM and then outline
NSM.

\begin{algorithm}[htb!]
\caption{Gillespie's direct method (DM)}
\begin{algorithmic}
  \STATE{\textit{Initialize:} Set the initial state $\fatx$ and
  compute all propensities $\omega_r(\fatx), r=1, \ldots,
  \Mreactions$. Also set $t=0$.}

  \WHILE{$t < T$}

  \STATE{Compute the sum $\lambda$ of all the propensities.}

  \STATE{Sample the next reaction time (by inversion), $\tau = -
  \log(\rand)/\lambda$. Here and in what follows, `$\rand$'
  conveniently denotes a uniformly distributed random number in
  $(0,1)$ which is different for each occurrence.}
  
  \STATE{Sample the next reaction event (by inversion); find $n$ such
  that \newline $\sum_{j=1}^{n-1} \omega_j(\fatx) < \lambda \, \rand
  \le \sum_{j=1}^n \omega_j(\fatx)$}

  \STATE{Update the state vector, $\fatx = \fatx+N_n$ and set
  $t=t+\tau$.}

  \ENDWHILE

\end{algorithmic}
\label{alg:SSADM}
\end{algorithm}

\begin{algorithm}[htb!]
\caption{Gibson and Bruck's next reaction method (NRM)}
\begin{algorithmic}
  \STATE{\textit{Initialize:} Set $t = 0$ and assign the initial
  number of molecules. Generate the dependency graph $G$. Compute the
  propensities $\omega_{r}(\fatx)$ and generate the corresponding
  \emph{absolute} waiting times $\tau_{r}$ for all reactions
  $r$. Store those values in a heap $H$.}

  \WHILE{$t < T$}

  \STATE{Remove the smallest time $\tau_{n} = H_{0}$ from the top of
  $H$, execute the $n$th reaction $\fatx := \fatx+N_{n}$ and set $t :=
  \tau_{n}$.}

  \FORALL{edges $n \to j$ in $G$}

  \IF{$j \not = n$}

  \STATE{Recompute the propensity $\omega_{j}$ and update
    the corresponding waiting time according to
\begin{align*}
  \tau_{j}^{\mathrm{new}} &= t+\left(\tau_{j}^{\mathrm{old}}-t\right)
  \frac{\omega_{j}^{\mathrm{old}}}{\omega_{j}^{\mathrm{new}}}.
\end{align*}}

  \ELSE[$j = n$]

  \STATE{Recompute the propensity $\omega_{n}$ and generate a new
  absolute time $\tau_{n}^{\mathrm{new}}$. Adjust the contents of $H$
  by replacing the old value of $\tau_{n}$ with the new one.}

  \ENDIF

  \ENDFOR

  \ENDWHILE
\end{algorithmic}
\label{alg:NRM}
\end{algorithm}

\begin{algorithm}[htb!]
\caption{The next subvolume method (NSM)}
\label{alg:NSM}
\begin{algorithmic}
  \STATE{\textit{Initialize:} Compute the sum $\sigma_i^r$ of all
  reaction rates $\omega_{ri}$ and the sum $\sigma_i^d$ of all
  diffusion rates $a_{ij}\fatx_{si}$ in all subvolumes $i=1,\ldots
  ,\Ncells$. Compute the time until the next event in each subvolume,
  $\tau_i = -\log(\rand)/(\sigma_i^r+\sigma_i^d)$, and store all times
  in a heap $H$.}

  \WHILE{$t < T$}

  \STATE{Select the next subvolume $\zeta_n$ where an event takes
  place by extracting the minimum $\tau_n$ from the top of $H$.}
  
  \STATE{Set $t=\tau_n.$} 

  \STATE{Determine if the event in $\zeta_n$ is a reaction or a
  diffusion event. Let it be a reaction if $(\sigma_n^r+\sigma_n^d) \,
  \rand < \sigma_n^r$, otherwise it is a diffusion event.}

  \IF{Reaction event}
  
  \STATE{Determine the reaction channel that fires. This is done by
  inversion of the distribution for the next reaction given $\tau_n$
  in the same manner as in Gillespie's direct method in
  Algorithm~\ref{alg:SSADM}.}

  \STATE{Update the state matrix using the (sparse) stoichiometric
    matrix $N$.}

  \STATE{Update $\sigma_n^r$ and $\sigma_n^d$ using the dependency
  graph $G$ to recalculate only affected reaction- and diffusion
  rates.}

  \ELSE[Diffusion event]

  \STATE{Determine which species $S_{ln}$ diffuses and subsequently,
  determine to which neighboring subvolume $\zeta_{n'}$. This is
  again done by inversion using a linear search in the corresponding
  column of $D$.}

  \STATE{Update the state: $S_{nl}=S_{nl}-1$, $S_{n'l}=S_{n'l}+1$.}
  
  \STATE{Update the reaction- and diffusion rates of subvolumes
  $\zeta_n$ and $\zeta_{n'}$ using G.}
  
  \ENDIF
  
  \STATE{Compute a new waiting time $\tau_n$ by drawing a new random
  number and add it to the heap $H$.}
  
  \ENDWHILE 
\end{algorithmic}
\end{algorithm}

